### 2017-04-06 paper/sexSelectionSpacing-ver1.tex
Editing estimation strategy section

As the discussion above shows there are three main
implications of the availability of sex-selective abortions
combined with son preference:
a higher probability that next child is a son, 
longer spacing to next birth with use of sex selection, 
and that the use of prenatal sex determination may change between 
one observed birth and the next.
All three must be incorporated into the estimation strategy
to provide unbiased and precise estimates of the use of sex-selective abortions.%

To understand son preference when sex selection is available,
we need an empirical model that can jointly estimate the sex of
children born and the spacing between birth.

The empirical model is a discrete time non-proportional competing risk hazard model with 
two exit states: either a boy or a girl is born.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}
A woman's reproductive life is divided into spells which each covers 
the period between births (from marriage to first birth for the first spell).
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$ and 
the spell continues until time $t_i$ when either a birth occurs or the spell is censored.
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.

There are two exit states: birth of a boy, $j=1$, or birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is
\begin{equation}
 h_{ijt} = \Pr (T_i=t, J_i=j \mid T_i \geq t; \mathbf{Z}_{it},\mathbf{X}_{i} ),
\end{equation}
where $T_i$ is a discrete random variable that captures when woman $i$'s birth occurs.
To ease presentation the indicator for spell number is suppressed.
The vectors of explanatory variable $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$ include 
information about various individual, household, and community characteristics 
discussed below.

@@

Specifying the model as a proportional hazard model, i.e.\ one where explanatory variables 
simply shift the hazard rates up or down independent of spell length, results in 
greater efficiency provided the proportionality assumption holds.
The problem is that the proportional hazard model does not allow covariates
to have different effects over time within a spell and ignoring differences in the shape 
of the hazard functions between different types of individuals can lead to substantial bias.
It is highly unlikely, even in the absence of prenatal sex determination, that the 
baseline hazards are the same across education levels, areas of residence, or sex 
composition of previous births.
The bias from the proportionality assumption is likely exacerbated by the 
introduction of prenatal sex determination for two reasons.
First, one of the main points of this paper is that use of sex-selective abortions 
affects birth spacing, and use of sex selection differs across groups.
Second, as discussed above, the use of sex selection may vary within a spell depending
on the length of the spell.

@@

The remaining explanatory variables, $\mathbf{X}$, enter proportionally.
% %
% \footnote{
% It is possible to incorporate time-variant characteristics of the explanatory
% variables in this set-up, but none of the explanatory variable change within a spell here.
% }
To further minimize any potential bias from assuming proportionality, estimations are 
done separately for different levels of mothers' education and for different time periods.
The exact specifications and the individual variables are described below.

@@

Direct interpretation of the estimated coefficients for this model is challenging because 
of the competing risk setup.
First, coefficients show the change in hazards relative to the base outcome, no birth, 
rather than simply the hazard of an event.
Second, a positive coefficient does not necessarily imply that an increase in a
variable's value increases the probability of the associated event because the 
probability of another event may increase even more \citep{thomas96}.

The model does, however, make it straightforward to calculate the 
predicted probabilities of having a boy and of having a girl for each $t$, conditional
on a set of explanatory variables and not having had a child before that period.
The predicted probability of having a boy in period $t$ for a given set of explanatory
variable values, $\mathbf{Z}_k$ and $\mathbf{X}_k$, is
\begin{equation}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})},
\label{eq:probability_boy}
\end{equation}
and the predicted probability of having a girl is
\begin{equation}
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}.
\label{eq:probability_girl}
\end{equation}
With these two probabilities is it easy to calculate, for each $t$, the estimated
percentage of children born that are boys, $\hat{Y}$, 
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100,
\label{eq:probability_son}
\end{equation}
and the associated confidence interval for given values of explanatory variables.%
\footnote{
Within each period $1-P(b_{t})-P(g_{t})$ is the probability of not having a birth in 
period $t$.
}

@@

Another important aspect of the hazard model is the survival curve, which shows the 
probability of not having had a birth yet by spell duration.
The survival curve for $t$ is 
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t 
\left( 
	1- \left(P(b_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) 
	+ P(g_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) \right) 
\right),
 \label{eq:survival}
\end{equation}
or equivalently
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t
\left(
\frac{ 1 }
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{ld}'\mathbf{Z}_{kd} + \beta_l'\mathbf{X}_{k})}
\right).
\end{equation}

[This need a much better explanation; the survival curve captures the spacing
between births and how long it takes; may need to combine with and without
sons to show differences in spacing over time; maybe it should also have
predicted average/median spacing with some confidence intervals to test]

The survival curves are presented for two reasons.
First, they show progression to the next birth and how quickly 
it occurs, if at all.
In the absence of sex selection, I expect most parities to show an 
``inverted s'' pattern, where there initially are relatively few births, 
followed by a substantial number of births over a 1 to 2 year period, and then
relatively few births thereafter.
As sex selection becomes more widely used the associated
longer spacing shows up by making the survival curve straighter,
indicating that some of the births that would originally have taken place
now take place later because of abortions.
The use of sex selection is clearly not the only factor that can change the shape of
the survival curves; factors such as the desired number of children and
use of contraceptives may also shape the shape.
To account for this it is best to compare survival curves for an individual
parity between women who are likely to use sex selection and women who are
not, for example because they already have one or more sons.

Secondly, the survival curves provide ``weighting'' for the associated percentage 
boys born.
The steeper the survival curve, the more weight should be assigned to a given
spell period because it is based on more births, 
whereas a period with a flat survival curve should be given little weight because the 
percentage boys is based on few births.
Hence, although they cannot by themselves show sex selection, the survival
curves are a crucial complement to the estimated sex ratios by duration.

>>>>

To improve on the literature's current approaches to understanding 
son preference, a new empirical method must be able to capture the 
three main implications of the availability of sex-selective abortions
combined with son preference:
a higher probability that the next child is a son, 
longer spacing to next birth with use of sex selection, 
and that the use of prenatal sex determination may change between 
one observed birth and the next.%
\footnote{
In addition, it should be able to capture the shorter spacing between
births observed in the absence of use or availability of prenatal 
sex determination.
}
This means that we need an empirical model that can jointly estimate 
the association between covariates and both the sex of children born 
and the spacing from last birth. 
My proposed model is a discrete time, non-proportional competing risk 
hazard model with two exit states: either a boy or a girl is born.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

I divide a woman's reproductive life into spells that each covers 
the period between births (from marriage to first birth for the first spell).
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$ and 
the spell continues until time $t_i$ when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: birth of a boy, $j=1$, or birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is
\begin{equation}
 h_{ijt} = \Pr (T_i=t, J_i=j \mid T_i \geq t; \mathbf{Z}_{it},\mathbf{X}_{i} ),
\end{equation}
where $T_i$ is a discrete random variable that captures when woman $i$'s birth occurs.
To ease presentation the indicator for spell number is suppressed.
The vectors of explanatory variable $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$ include 
information about various individual, household, and community characteristics 
discussed below.

@@

[TK I am not very satisfied with this explanation of the problems with
proportionality and the associated bias; non-proportionality is required
to capture changes in use of sex selection]

In principle, specifying the model as a proportional hazard model, i.e.\ one 
where covariates simply shift the hazard rates up or down independent of 
spell length, is more efficient, but only provided that the proportionality 
assumption holds.
If the proportionality assumption does not hold, however, the result is
a potentially substantial bias in estimates.
The problem is that a proportional hazard model does not allow covariates
to have different effects at different times within a spell and therefore
cannot capture differences in the shape of the hazard functions between 
different groups. 
It is highly unlikely---even in the absence of prenatal sex determination---that 
the baseline hazards are the same across education levels, areas of residence, 
or sex composition of previous births.
Any bias from the proportionality assumption is likely exacerbated by the 
introduction of prenatal sex determination for two reasons.
First, one of the main points of this paper is that use of sex-selective 
abortions affects birth spacing, and use of sex selection differs across groups.
Second, as discussed above, the use of sex selection may vary within a spell,
depending on the length of the spell.

@@

The remaining explanatory variables, $\mathbf{X}$, enter proportionally.
To further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and for different 
time periods.
The exact specifications and the individual variables are described below.

@@

The main downside of my approach is that direct interpretation of the estimated 
coefficients for this model is challenging because of the competing risk setup.
First, coefficients show the change in hazards relative to the base outcome, 
no birth, rather than simply the hazard of an event.
Second, a positive coefficient does not necessarily imply that an increase in a
variable's value increases the probability of the associated event because the 
probability of another event may increase even more \citep{thomas96}.

It is, however, straightforward to calculate the predicted probabilities of 
having a boy and of having a girl for each $t$ within a spell, conditional on 
a set of explanatory variables and not having had a child before that period.
The predicted probability of having a boy in period $t$ for a given set of 
explanatory variable values, $\mathbf{Z}_k$ and $\mathbf{X}_k$, is
\begin{equation}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})},
\label{eq:probability_boy}
\end{equation}
and the predicted probability of having a girl is
\begin{equation}
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}.
\label{eq:probability_girl}
\end{equation}
With these two probabilities is it easy to calculate, for each $t$, the estimated
percentage of children born that are boys, $\hat{Y}$, 
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100,
\label{eq:probability_son}
\end{equation}
together with the associated confidence interval for given values of explanatory 
variables.%
\footnote{
Within each period $1-P(b_{t})-P(g_{t})$ is the probability of not having a birth in 
period $t$.
}

@@

[TK  may need to combine with and without sons to show differences in spacing 
over time; maybe it should also have predicted average/median spacing with some 
confidence intervals to test]

The other important part of the model is the spacing between births.
Spacing is captured by the survival curve, which shows the probability of not 
having had a birth yet by spell duration.
The survival curve at time $t$ is 
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t 
\left( 
	1- \left(P(b_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) 
	+ P(g_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) \right) 
\right),
 \label{eq:survival}
\end{equation}
or equivalently
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t
\left(
\frac{ 1 }
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{ld}'\mathbf{Z}_{kd} + \beta_l'\mathbf{X}_{k})}
\right).
\end{equation}

In the absence of sex selection, I expect most parities to show an 
``inverted s'' pattern, where there initially are relatively few births, 
followed by a substantial number of births over a 1 to 2 year period, and then
relatively few births thereafter.
As sex selection becomes more widely used the associated
longer spacing shows up by making the survival curve straighter,
indicating that some of the births that would originally have taken place
now take place later because of abortions.
The use of sex selection is clearly not the only factor that can change the shape of
the survival curves; factors such as the desired number of children and
use of contraceptives may also shape the shape.
To account for this it is best to compare survival curves for an individual
parity between women who are likely to use sex selection and women who are
not, for example because they already have one or more sons.

Furthermore, the survival curves also provide ``weighting'' for the 
associated percentage boys born.
The steeper the survival curve, the more weight should be assigned to a given
spell period because it is based on more births, 
whereas a period with a flat survival curve should be given little weight because the 
percentage boys is based on few births.
Hence, although they cannot by themselves show sex selection, the survival
curves are a crucial complement to the estimated sex ratios by duration.
